\chapter*{Introduction}

In order to properly implement the design decisions made for the PLeTs v4 project, it is necessary that all professionals involved be aware of certain mechanisms, both theoretical and practical, that will be used in its design and development. This document is aimed to serve both as a basis for the study and explanation of these techniques and as a guide for the appropriate representation and implementation of them in PLeTs v4.

In Chapter \ref{ch:mechanisms} we will present some variability management mechanisms, as well as explain which ones to use, when each is appropriate and why some should not be used. First a basic theory of the concept is presented, based on \cite{APEL:2013}. The mechanisms of using Parameters, Design Patterns, Frameworks, Components and Services, and Preprocessors are presented. Then, the idea behind the particular variability mechanisms of PLeTs v4 are presented, with their in-depth descriptions being left to \ref{ch:patterns}.

In Chapter \ref{ch:patterns} we will present the basics of the software design patterns and architectural patterns used in PLeTs v4, with a focus on explaining their goals and implementation. First the Factory Method design pattern will be presented \cite{LASER:2015} alongside the data structure catalogue system implemented in PLeTs v3. We then briefly expose the basic concepts of Software Connectors \cite{TAYLOR:2009}, with a particular emphasis on Linkage and Arbitrator connectors. With a basis on these concepts, we then proceed to present the architectural pattern proposed for the composition of products in PLeTs v4, which is largely based on the Distributed Objects architectural style \cite{CORBA:2012}.

Chapter \ref{ch:modules} is meant to serve as a guide to the components designed within PLeTs v4 and their specifications. This chapter is focused on the architectural characteristics of these components, and is not meant to serve as a guide to the software testing methods applied in PLeTs.

Chapter \ref{ch:architecture} is reserved for track keeping of the evolution of the PLeTs v4 architecture project. It is my intention that this chapter be constantly updated and kept as a snapshot of the current state of the project. Any impreciseness or incorrectness in this chapter is to be taken as architectural erosion, and must be corrected as soon as possible.

Appendix \ref{ch:compilers} is reserved to the study of basic concepts of Formal Languages and Compilers, and is left as an aid for the development of parsers and script generators for PLeTs v4.

Should this document at any point prove too dense, complicated, unclear or in any way lacking, I ask that you inform me (marcelo.laser@gmail.com) as quickly as possible so that the problem may be resolved and the document may be kept in the best possible shape. Please bear in mind that this document is to serve as the primary guide of the PLeTs v4 infrastructure, and therefore it aims to be as clear and precise as possible.

Marcelo Schmitt Laser
